Resolvable Ambiguity

### 2019-07-08 paper/paper.tex
Introduce the notion of an abstract parser

This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to some more standard concepts in formal languages.

Normally, we define a language as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$, but here we additionally require the existence of a \emph{meaning} of each word, in some sense. We call this meaning an \emph{interpretation} of a word. Note that a single word may have multiple interpretations. As such, we define an \emph{interpreted language} as:

\begin{definition}
  An interpreted language, written $L(\T, T, \parse)$, consists of
\begin{itemize}
\item an alphabet $\T$,
\item a set of interpretations $T$, and
\item a function $\parse : \T^{*} -> 2^T$ that relates words to their interpretations, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
\end{itemize}

  A word $w \in \T^{*}$ is in the language $L(\T, T, \parse)$, written
  $w \in L(\T, T, \parse)$ iff $\parse(w) \neq \emptyset$.
\end{definition}

\noindent For example, consider a simple arithmetic language without precedence and parentheses, where the interpretations are abstract syntax trees. In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two ASTs shown in Fig.~\ref{fig:arith-example}.

\begin{figure}[t]
\begin{tabular}{cc}
  \Tree [.{$+$}
    1
    [.{$\cdot$}
      2
      3 ] ] &
  \Tree [.{$\cdot$}
    [.{$+$}
      1
      2 ]
    3 ] \\
\end{tabular}
\caption{The two allowable interpretations of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
\label{fig:arith-example}
\end{figure}

@@

\begin{definition}
  A word $w \in L(\T, T, \parse)$ is ambiguous, written $\amb_L(w)$, iff

@@

\noindent
Note that the above definition implies that a word $w \in L(\T, T,
\parse)$ is not ambiguous (or unambiguous) if $\exists t \in
T.\ \parse(w) = \{t\}$.

We can connect the above definition of an interpreted language to the
classical definition of a language as follows:

\begin{itemize}
\item Given an interpreted language $L(\T, T, \parse)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

\noindent A \emph{resolvably ambiguous} word is a word where all its interpretations can be written in an unambiguous way, formally:

\begin{definition}\label{def:resolvable-word}
  A word $w \in L(\T, T, \parse)$ is resolvably ambiguous, written $\rho_L(w)$, iff

@@

The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously interpreted as the first and second tree, respectively.

Additionally, we define a \emph{language} to be resolvably ambiguous if all its words are resolvably ambiguous, formally:

\begin{definition}\label{def:resolvable-language}
  An interpreted language $L(\T, T, \parse)$ is resolvably ambiguous, written $\rho(L)$, iff

  $\forall w \in L(\T, T, \parse).\ \rho_L(w)$.
\end{definition}

@@

\begin{itemize}
\item An unambigous word $w$ is trivially resolvably ambiguous, since its only interpretation $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$). The set of resolvably ambiguous words is thus a superset of the unambiguous words.
\item If a given interpretation $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous. In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
\end{itemize}

\noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$. Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual interpretation. Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace. In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the interpretation.

@@

\begin{description}
\item[Static resolvability.] Given a language $L(\T, T, \parse)$, determine whether $\rho(L)$.
\item[Dynamic resolvability.] Given a word $w \in L(\T, T, \parse)$, determine whether $\rho_L(w)$.
\end{description}

>>>>

This section introduces our definition of \emph{resolvable ambiguity}, and then relates it to standard concepts in formal languages.

A formal language is defined as a set of words, i.e., a subset of $\T^{*}$ for some alphabet $\T$. For defining \emph{resolvable ambiguity} we additionally have to consider the results of parsing words. In order to stay as general as possible, we first define the notion of an \emph{abstract parser}.

\begin{definition}
  An abstract parser $P$ is a triple $(\T, T, \parse)$ consisting of
\begin{itemize}
\item an alphabet $\T$,
\item a set of parse trees $T$, and
\item a function $\parse : \T^{*} -> 2^T$ that relates words to their parse trees, where $2^T$ denotes the powerset of $T$. Additionally, we require that $T = \bigcup_{w \in \T^{*}} \parse(w)$.
\end{itemize}
\end{definition}

This notion of an abstract parser enables the introduction of a particular class of formal languages that we will use throughout the rest of the paper; we call members of this class \emph{parse languages:}

\begin{definition}
  Given a word $w \in \T^{*}$ and an abstract parser $P = (\T, T,
  \parse)$, the set of words contained in the \emph{parse language}
  $L(P)$ is defined as follows:

  $w \in L(P)$ iff $\parse(w) \neq \emptyset$.
\end{definition}

\noindent For example, consider a simple arithmetic language without precedence and parentheses. In such a language, $\parse(1 + 2 \cdot 3)$ would produce a set containing the two parse trees shown in Fig.~\ref{fig:arith-example}.

\begin{figure}[t]
\begin{tabular}{cc}
  \Tree [.{$+$}
    1
    [.{$\cdot$}
      2
      3 ] ] &
  \Tree [.{$\cdot$}
    [.{$+$}
      1
      2 ]
    3 ] \\
\end{tabular}
\caption{The two allowable parse trees of $1 + 2 \cdot 3$ in an arithmetic language without precedence.}
\label{fig:arith-example}
\end{figure}

@@

\begin{definition}
  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$
  is ambiguous, written $\amb_P(w)$, iff

@@

\noindent
Note that the above definition implies that a word $w \in L(P)$, where
$P = (\T, T, \parse)$, is not ambiguous, or \emph{unambiguous}, if
$\exists t \in T.\ \parse(w) = \{t\}$.

We can connect the above definition of a parse language to the
classical definition of a (formal) language as follows:

\begin{itemize}
\item Given a parse language $L(P)$, the corresponding classical language (i.e., set of words) is given by $\{ w \mid \parse(w) \neq \emptyset \}$.
\item If we select a $\parse$ function that relates words to their leftmost derivations in a given context-free grammar, then our definition of ambiguity corresponds exactly to the classical definition of ambiguity.
\end{itemize}

\noindent A \emph{resolvably ambiguous} word is a word where all its parse trees can be written in an unambiguous way, formally:

\begin{definition}\label{def:resolvable-word}
  Given an abstract parser $P = (\T, T, \parse)$, a word $w \in L(P)$ is \emph{resolvably ambiguous}, written $\rho_P(w)$, iff

@@

The example in Fig.~\ref{fig:arith-example} is not resolvably ambiguous, since neither tree can be written in any other way. However, if we add grouping parentheses to the language we find that $1 + (2 \cdot 3)$ and $(1 + 2) \cdot 3$ are unambiguously parsed as the first and second tree, respectively.

Additionally, we define an \emph{abstract parser} to be resolvably ambiguous if all words in its corresponding parse language $L(P)$ are resolvably ambiguous, formally:

\begin{definition}\label{def:resolvable-language}
  An abstract parser $P$ is resolvably ambiguous, written $\rho(P)$, iff

  $\forall w \in L(P).\ \rho_P(w)$.
\end{definition}

@@

\begin{itemize}
\item An unambigous word $w$ is trivially resolvably ambiguous, since its only parse tree $t$ can be written unambiguously with $w$ itself ($\parse(w) = \{t\}$). The set of resolvably ambiguous words is thus a superset of the unambiguous words.
\item If a given parse tree $t$ has only one word $w$ such that $t \in \parse(w)$, then $w$ is resolvably ambiguous iff it is unambiguous. In general, $\forall t \in T.\ \lvert\{w \mid t \in \parse(w)\}\rvert \leq 1$ implies that the set of resolvably ambiguous words is exactly the set of unambiguous words.
\end{itemize}

\noindent The second point suggests that resolvable ambiguity is only an interesting property if an element of $T$ does not uniquely identify an element of $\T^{*}$. Intuitively, this only happens if $\parse$ discards some information present in its argument when constructing an individual parse tree. Fortunately, this is generally true for parsing in commonly used programming languages; they tend to discard, e.g., grouping parentheses and whitespace. In general, whatever information $\parse$ discards can be used by an end-user to disambiguate an ambiguity encountered at parse-time, without changing the parse tree.

@@

\begin{description}
\item[Static resolvability.] Given an abstract parser $P$, determine whether $\rho(P)$.
\item[Dynamic resolvability.] Given a word $w \in L(P)$ for some abstract parser $P$, determine whether $\rho_P(w)$.
\end{description}
